---
title: "Numerical MLE and Likelihood-Based Inference"
subtitle: "Exponential and Gamma models for whale diving times"
author: "George Oikonomidis"
editor: visual
---

## Aim

This case study connects numerical optimization with statistical inference. Maximum-likelihood estimates are obtained numerically, checked against analytical results where available, and used to construct Wald and likelihood-ratio confidence intervals.

```{r}
#| label: setup
options(digits = 5)

whales <- scan("data/whales.txt", quiet = TRUE)
stopifnot(length(whales) > 0, all(whales > 0))

summary(whales)
```

## Exponential model

Let $Y_1,\ldots,Y_n$ be independent exponential observations with **rate** $\lambda>0$:

$$f(y;\lambda)=\lambda e^{-\lambda y}.$$

The log-likelihood is

$$\ell(\lambda)=n\log\lambda-\lambda\sum_{i=1}^n y_i.$$

```{r}
loglik_exponential <- function(lambda, y) {
  if (lambda <= 0) return(-Inf)
  length(y) * log(lambda) - lambda * sum(y)
}

neg_loglik_exponential <- function(lambda, y) {
  -loglik_exponential(lambda, y)
}

lambda_grid <- seq(0.2, 3.5, length.out = 400)
plot(
  lambda_grid,
  vapply(lambda_grid, loglik_exponential, numeric(1), y = whales),
  type = "l", lwd = 2,
  xlab = expression(lambda), ylab = "log-likelihood",
  main = "Exponential log-likelihood"
)
```

## Numerical MLE and analytical verification

The optimization is bounded away from zero and requests a numerical Hessian.

```{r}
fit_exponential <- optim(
  par = 1,
  fn = neg_loglik_exponential,
  y = whales,
  method = "L-BFGS-B",
  lower = 1e-8,
  hessian = TRUE
)

if (fit_exponential$convergence != 0) {
  stop("The exponential optimization did not converge.")
}

lambda_mle_numerical <- unname(fit_exponential$par)
lambda_mle_analytical <- 1 / mean(whales)

data.frame(
  method = c("Numerical MLE", "Analytical MLE"),
  rate = c(lambda_mle_numerical, lambda_mle_analytical),
  mean_scale = 1 / c(lambda_mle_numerical, lambda_mle_analytical)
)
```

The distinction between rate $\lambda$ and mean/scale $1/\lambda$ is explicit because `rexp()` and `dexp()` use `rate` by default.

## Hessian-based standard error and Wald interval

The observed information is the Hessian of the negative log-likelihood at the MLE. Its inverse approximates the estimator's variance.

```{r}
observed_information <- fit_exponential$hessian[1, 1]
se_lambda <- sqrt(1 / observed_information)

wald_ci <- lambda_mle_numerical + qnorm(c(0.025, 0.975)) * se_lambda

data.frame(
  estimate = lambda_mle_numerical,
  standard_error = se_lambda,
  lower_95 = wald_ci[1],
  upper_95 = wald_ci[2]
)
```

## Likelihood-ratio confidence interval

The approximate 95% likelihood-ratio region contains parameter values satisfying

$$2\{\ell(\widehat\lambda)-\ell(\lambda)\}\leq\chi^2_{1,0.95}.$$

```{r}
lr_distance <- function(lambda, y, lambda_hat, level = 0.95) {
  2 * (loglik_exponential(lambda_hat, y) - loglik_exponential(lambda, y)) -
    qchisq(level, df = 1)
}

lr_lower <- uniroot(
  lr_distance,
  interval = c(1e-6, lambda_mle_numerical),
  y = whales,
  lambda_hat = lambda_mle_numerical
)$root

lr_upper <- uniroot(
  lr_distance,
  interval = c(lambda_mle_numerical, 5),
  y = whales,
  lambda_hat = lambda_mle_numerical
)$root

interval_comparison <- data.frame(
  method = c("Wald", "Likelihood ratio"),
  lower = c(wald_ci[1], lr_lower),
  upper = c(wald_ci[2], lr_upper)
)

interval_comparison
```

```{r}
loglik_values <- vapply(
  lambda_grid,
  loglik_exponential,
  numeric(1),
  y = whales
)

plot(lambda_grid, loglik_values, type = "l", lwd = 2,
     xlab = expression(lambda), ylab = "log-likelihood",
     main = "Likelihood-ratio interval")
abline(v = c(lr_lower, lambda_mle_numerical, lr_upper),
       lty = c(2, 1, 2), col = c("steelblue", "firebrick", "steelblue"))
abline(h = loglik_exponential(lambda_mle_numerical, whales) -
         qchisq(0.95, 1) / 2,
       lty = 3)
```

## Gamma extension

The Gamma model adds a shape parameter $\alpha$ and uses a rate parameter $\beta$:

$$f(y;\alpha,\beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}
y^{\alpha-1}e^{-\beta y}.$$

Optimizing on the log-parameter scale guarantees $\alpha,\beta>0$.

```{r}
neg_loglik_gamma <- function(log_parameters, y) {
  shape <- exp(log_parameters[1])
  rate <- exp(log_parameters[2])
  -sum(dgamma(y, shape = shape, rate = rate, log = TRUE))
}

initial_shape <- 1
initial_rate <- initial_shape / mean(whales)

fit_gamma <- optim(
  par = log(c(initial_shape, initial_rate)),
  fn = neg_loglik_gamma,
  y = whales,
  method = "BFGS",
  hessian = TRUE
)

if (fit_gamma$convergence != 0) {
  stop("The Gamma optimization did not converge.")
}

gamma_parameters <- exp(fit_gamma$par)
names(gamma_parameters) <- c("shape", "rate")

model_comparison <- data.frame(
  model = c("Exponential", "Gamma"),
  parameters = c(1, 2),
  log_likelihood = c(-fit_exponential$value, -fit_gamma$value)
)
model_comparison$AIC <- with(
  model_comparison,
  2 * parameters - 2 * log_likelihood
)

gamma_parameters
model_comparison
```

The Gamma model should only be preferred if the additional flexibility produces a meaningful improvement after accounting for its extra parameter. A lower AIC is evidence in that direction, not a substitute for model diagnostics.

## Profile likelihood for a parameter

For a parameter of interest $\psi$ and nuisance parameter $\eta$, the profile log-likelihood is

$$\ell_p(\psi)=\max_{\eta}\ell(\psi,\eta).$$

Profiling is different from fixing the nuisance parameter at its joint MLE: it is re-estimated at every candidate value of $\psi$.

For the Gamma model, if the shape $\alpha$ is fixed, maximization over the rate has the analytical solution

$$\widehat\beta(\alpha)=\frac{n\alpha}{\sum_i y_i}
=\frac{\alpha}{\overline y}.$$

```{r}
gamma_loglik <- function(shape, rate, y) {
  if (shape <= 0 || rate <= 0) return(-Inf)
  sum(dgamma(y, shape = shape, rate = rate, log = TRUE))
}

profile_shape <- function(shape, y) {
  if (shape <= 0) return(-Inf)
  conditional_rate <- shape / mean(y)
  gamma_loglik(shape, conditional_rate, y)
}

shape_hat <- unname(gamma_parameters["shape"])
profile_cutoff <- -fit_gamma$value - qchisq(0.95, df = 1) / 2

shape_root <- function(shape) {
  profile_shape(shape, whales) - profile_cutoff
}

shape_profile_ci <- c(
  lower = uniroot(shape_root, c(0.05, shape_hat))$root,
  upper = uniroot(shape_root, c(shape_hat, 10))$root
)

data.frame(
  parameter = "Gamma shape",
  estimate = shape_hat,
  lower_95 = shape_profile_ci["lower"],
  upper_95 = shape_profile_ci["upper"]
)
```

The likelihood-ratio profile interval contains values satisfying

$$2\{\ell_p(\widehat\alpha)-\ell_p(\alpha)\}
\leq \chi^2_{1,0.95}.$$

## Profile likelihood for a function of parameters

Now let the parameter of interest be the Gamma mean

$$\mu=\frac{\alpha}{\beta}.$$

For a fixed value of $\mu$, reparameterize $\beta=\alpha/\mu$ and maximize over the nuisance shape $\alpha$. This profiles a **function of two original parameters** without relying on a delta-method approximation.

```{r}
profile_mean <- function(mu, y) {
  if (mu <= 0) return(-Inf)

  conditional_fit <- optimize(
    function(log_shape) {
      shape <- exp(log_shape)
      rate <- shape / mu
      -gamma_loglik(shape, rate, y)
    },
    interval = log(c(0.05, 50))
  )

  -conditional_fit$objective
}

mean_hat <- unname(gamma_parameters["shape"] / gamma_parameters["rate"])

mean_root <- function(mu) {
  profile_mean(mu, whales) - profile_cutoff
}

mean_profile_ci <- c(
  lower = uniroot(mean_root, c(0.05, mean_hat))$root,
  upper = uniroot(mean_root, c(mean_hat, 2))$root
)

data.frame(
  function_of_parameters = "Gamma mean = shape / rate",
  estimate = mean_hat,
  lower_95 = mean_profile_ci["lower"],
  upper_95 = mean_profile_ci["upper"]
)
```

```{r}
shape_grid <- seq(shape_profile_ci["lower"] * 0.8,
                  shape_profile_ci["upper"] * 1.2,
                  length.out = 300)
mean_grid <- seq(mean_profile_ci["lower"] * 0.8,
                 mean_profile_ci["upper"] * 1.2,
                 length.out = 300)

par(mfrow = c(1, 2))
plot(
  shape_grid,
  vapply(shape_grid, profile_shape, numeric(1), y = whales),
  type = "l", lwd = 2,
  xlab = expression(alpha), ylab = "Profile log-likelihood",
  main = "Profile for Gamma shape"
)
abline(h = profile_cutoff, lty = 3)
abline(v = c(shape_profile_ci["lower"], shape_hat,
             shape_profile_ci["upper"]),
       lty = c(2, 1, 2), col = c("steelblue", "firebrick", "steelblue"))

plot(
  mean_grid,
  vapply(mean_grid, profile_mean, numeric(1), y = whales),
  type = "l", lwd = 2,
  xlab = expression(mu), ylab = "Profile log-likelihood",
  main = "Profile for Gamma mean"
)
abline(h = profile_cutoff, lty = 3)
abline(v = c(mean_profile_ci["lower"], mean_hat,
             mean_profile_ci["upper"]),
       lty = c(2, 1, 2), col = c("steelblue", "firebrick", "steelblue"))
par(mfrow = c(1, 1))
```

Profile likelihood automatically reflects asymmetry and the dependence between parameters. It is especially useful when a Wald interval is distorted by a nonlinear transformation or when the likelihood is not close to quadratic.

## Key checks

- Use log-densities rather than products of densities.
- Encode positivity through bounds or a log transformation.
- Check the optimizer's convergence code.
- Confirm that the observed information is positive before reporting standard errors.
- Compare numerical results with analytical solutions whenever one exists.
