---
title: "Likelihood, Bootstrap, and Rejection Sampling"
subtitle: "Three computational statistics exercises"
author: "George Oikonomidis"
editor: visual
---

```{r}
#| label: setup
set.seed(2026)
options(digits = 5)
```

## Exercise 1: inverse-Gamma MLE for whale diving times

The diving times are modelled with the inverse-Gamma density

$$f(x;\alpha,\beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}
x^{-(\alpha+1)}e^{-\beta/x},
\qquad x>0.$$

The log-likelihood is

$$\ell(\alpha,\beta)=n\alpha\log\beta-n\log\Gamma(\alpha)
-(\alpha+1)\sum_i\log x_i-\beta\sum_i\frac{1}{x_i}.$$

```{r}
whales <- scan("data/whales.txt", quiet = TRUE)

loglik_inverse_gamma <- function(alpha, beta, x) {
  if (alpha <= 0 || beta <= 0) return(-Inf)

  n <- length(x)
  n * alpha * log(beta) - n * lgamma(alpha) -
    (alpha + 1) * sum(log(x)) - beta * sum(1 / x)
}

c(
  minimum = min(whales),
  maximum = max(whales),
  observations = length(whales)
)
```

Assume $\beta=1$. The score and its derivative are

$$\frac{\partial\ell}{\partial\alpha}
=n\log\beta-n\,\psi(\alpha)-\sum_i\log x_i,$$

$$\frac{\partial^2\ell}{\partial\alpha^2}
=-n\,\psi_1(\alpha),$$

where $\psi$ and $\psi_1$ are the digamma and trigamma functions.

```{r}
newton_inverse_gamma_shape <- function(x, beta = 1, start = 2,
                                       tolerance = 1e-10,
                                       max_iter = 100) {
  alpha <- start
  trace <- data.frame(iteration = 0, alpha = alpha)

  for (iteration in seq_len(max_iter)) {
    score <- length(x) * log(beta) -
      length(x) * digamma(alpha) - sum(log(x))
    hessian <- -length(x) * trigamma(alpha)
    step_fraction <- 1
    candidate <- alpha - step_fraction * score / hessian

    while ((!is.finite(candidate) || candidate <= 0) &&
           step_fraction > 2^-20) {
      step_fraction <- step_fraction / 2
      candidate <- alpha - step_fraction * score / hessian
    }

    if (!is.finite(candidate) || candidate <= 0) {
      stop("Newton-Raphson could not find an admissible step.")
    }

    trace <- rbind(
      trace,
      data.frame(iteration = iteration, alpha = candidate)
    )

    if (abs(candidate - alpha) <= tolerance * (1 + abs(alpha))) {
      return(list(estimate = candidate, converged = TRUE, trace = trace))
    }

    alpha <- candidate
  }

  list(estimate = alpha, converged = FALSE, trace = trace)
}

shape_fit <- newton_inverse_gamma_shape(whales, beta = 1, start = 2)

shape_check <- optimize(
  function(alpha) loglik_inverse_gamma(alpha, beta = 1, x = whales),
  interval = c(0.1, 10),
  maximum = TRUE
)

data.frame(
  method = c("Newton-Raphson", "optimize"),
  alpha_hat = c(shape_fit$estimate, shape_check$maximum),
  log_likelihood = c(
    loglik_inverse_gamma(shape_fit$estimate, 1, whales),
    shape_check$objective
  )
)
```

No elementary closed form exists for the inverse of the digamma equation, so a numerical root finder or optimizer is genuinely required for $\widehat\alpha$.

## Exercise 2: bootstrap bias and standard error of an exponential MLE

For $X_i\sim\operatorname{Exp}(\theta)$ with **rate** $\theta$,

$$\widehat\theta=\frac{n}{\sum_iX_i}=\frac{1}{\overline X}.$$

Generate samples from the model, estimate the bootstrap bias and standard error of $\widehat\theta$, and compare them with the theoretical values

$$\operatorname{Bias}(\widehat\theta)=\frac{\theta}{n-1},
\qquad
\operatorname{SE}(\widehat\theta)=
\frac{n\theta}{(n-1)\sqrt{n-2}}.$$

```{r}
bootstrap_exponential_rate <- function(x, B = 1000) {
  theta_hat <- 1 / mean(x)
  bootstrap_estimates <- replicate(
    B,
    1 / mean(sample(x, replace = TRUE))
  )

  c(
    estimate = theta_hat,
    bootstrap_bias = mean(bootstrap_estimates) - theta_hat,
    bootstrap_se = sd(bootstrap_estimates)
  )
}

n <- 25
theta <- 1
outer_replications <- 20

bootstrap_study <- t(replicate(
  outer_replications,
  bootstrap_exponential_rate(
    rexp(n, rate = theta),
    B = 1000
  )
))

comparison <- data.frame(
  quantity = c("Bias", "Standard error"),
  average_bootstrap_estimate = c(
    mean(bootstrap_study[, "bootstrap_bias"]),
    mean(bootstrap_study[, "bootstrap_se"])
  ),
  theoretical_value = c(
    theta / (n - 1),
    n * theta / ((n - 1) * sqrt(n - 2))
  )
)

comparison
```

With only 20 outer replications, the comparison itself has visible Monte Carlo error. The exercise is useful precisely because it separates the bootstrap approximation inside each sample from repeated-sampling performance across samples.

## Exercise 3: Gamma generation by rejection sampling

For a Gamma target with shape $\alpha=3$ and rate $\beta=2$, use an exponential proposal with rate $\lambda=1$. The ratio is maximized at

$$\widehat x=\frac{\alpha-1}{\beta-\lambda}=2.$$

```{r}
gamma_rejection_sample <- function(n, alpha = 3, beta = 2, lambda = 1) {
  stopifnot(alpha > 1, lambda > 0, lambda < beta)
  x_hat <- (alpha - 1) / (beta - lambda)
  result <- numeric(n)
  attempts <- 0L
  accepted <- 0L

  while (accepted < n) {
    V <- rexp(1, rate = lambda)
    W <- runif(1)
    acceptance_probability <-
      (V / x_hat)^(alpha - 1) *
      exp(-(beta - lambda) * (V - x_hat))

    attempts <- attempts + 1L
    if (W <= acceptance_probability) {
      accepted <- accepted + 1L
      result[accepted] <- V
    }
  }

  list(sample = result, acceptance_rate = n / attempts)
}

gamma_sample <- gamma_rejection_sample(1000)
gamma_sample$acceptance_rate
```

```{r}
hist(gamma_sample$sample, probability = TRUE, breaks = 30,
     col = "grey85", border = "white",
     main = "Gamma rejection sampler",
     xlab = "x")
curve(dgamma(x, shape = 3, rate = 2), add = TRUE,
      col = "firebrick", lwd = 2)
```
