---
title: "Gamma Rate Estimation"
subtitle: "Newton–Raphson and closed-form verification"
author: "George Oikonomidis"
editor: visual
---

```{r}
#| label: setup
options(digits = 5)
```

## Exercise

The recorded delivery times are modelled with a Gamma distribution having fixed shape $\alpha=2.5$ and unknown **rate** $\beta$. Write the log-likelihood, obtain the MLE numerically with Newton–Raphson, and verify whether a closed-form solution exists.

The log-likelihood is

$$\ell(\beta)=n\alpha\log\beta-n\log\Gamma(\alpha)
+(\alpha-1)\sum_i\log x_i-
\beta\sum_i x_i.$$

```{r}
delivery_times <- c(
  2.10, 3.40, 4.25, 5.60, 6.40, 7.30, 8.50, 8.75,
  8.90, 9.50, 9.75, 10.00, 10.40, 10.40, 16.00, 19.00
)

gamma_loglik_rate <- function(beta, alpha, x) {
  if (beta <= 0) return(-Inf)

  length(x) * alpha * log(beta) - length(x) * lgamma(alpha) +
    (alpha - 1) * sum(log(x)) - beta * sum(x)
}

alpha_fixed <- 2.5

c(
  observations = length(delivery_times),
  sample_mean = mean(delivery_times),
  loglik_at_beta_1 = gamma_loglik_rate(
    beta = 1,
    alpha = alpha_fixed,
    x = delivery_times
  )
)
```

### Newton–Raphson implementation

The score and Hessian are

$$\ell'(\beta)=\frac{n\alpha}{\beta}-\sum_i x_i,
\qquad
\ell''(\beta)=-\frac{n\alpha}{\beta^2}.$$

```{r}
newton_gamma_rate <- function(start, alpha, x,
                              tolerance = 1e-12,
                              max_iter = 100) {
  beta <- start
  trace <- data.frame(iteration = 0, beta = beta)

  for (iteration in seq_len(max_iter)) {
    score <- length(x) * alpha / beta - sum(x)
    hessian <- -length(x) * alpha / beta^2
    candidate <- beta - score / hessian

    if (!is.finite(candidate) || candidate <= 0) {
      stop("Newton step left the admissible domain beta > 0.")
    }

    trace <- rbind(
      trace,
      data.frame(iteration = iteration, beta = candidate)
    )

    if (abs(candidate - beta) <= tolerance * (1 + abs(beta))) {
      return(list(estimate = candidate, converged = TRUE, trace = trace))
    }

    beta <- candidate
  }

  list(estimate = beta, converged = FALSE, trace = trace)
}

beta_newton <- newton_gamma_rate(
  start = 0.1,
  alpha = alpha_fixed,
  x = delivery_times
)

beta_newton$trace
```

### Closed-form verification

The score equation can also be solved directly:

$$\widehat\beta=\frac{n\alpha}{\sum_i x_i}.$$

```{r}
beta_closed_form <- length(delivery_times) * alpha_fixed /
  sum(delivery_times)

data.frame(
  method = c("Newton-Raphson", "Closed-form MLE"),
  beta_hat = c(beta_newton$estimate, beta_closed_form),
  log_likelihood = c(
    gamma_loglik_rate(beta_newton$estimate, alpha_fixed, delivery_times),
    gamma_loglik_rate(beta_closed_form, alpha_fixed, delivery_times)
  )
)
```

The numerical method remains a useful computational check, but it is not required for this particular one-parameter problem.
