---
title: "Jackknife, BCa, and Regression Bootstrap"
subtitle: "Bias, influence, confidence intervals, and two regression resampling schemes"
author: "George Oikonomidis"
editor: visual
---

## Aim

This notebook extends ordinary bootstrap inference with the jackknife, BCa confidence intervals, influence diagnostics, and bootstrap methods for linear regression.

```{r}
#| label: setup
set.seed(2026)
options(digits = 5)

library(boot)
```

## Leave-one-out jackknife

For an estimator $\widehat\theta=T(x_1,\ldots,x_n)$, the $i$th jackknife replicate removes observation $i$:

$$\widehat\theta_{(i)}=T(x_1,\ldots,x_{i-1},x_{i+1},\ldots,x_n).$$

The jackknife estimates bias and standard error by

$$\widehat{\operatorname{Bias}}_{\mathrm{jack}}
=(n-1)(\overline\theta_{(\cdot)}-\widehat\theta),$$

$$\widehat{\operatorname{SE}}_{\mathrm{jack}}
=\sqrt{\frac{n-1}{n}\sum_{i=1}^n
(\widehat\theta_{(i)}-\overline\theta_{(\cdot)})^2}.$$

```{r}
jackknife_summary <- function(x, statistic) {
  n <- length(x)
  estimate <- statistic(x)
  replicates <- numeric(n)

  for (i in seq_len(n)) {
    replicates[i] <- statistic(x[-i])
  }

  replicate_mean <- mean(replicates)

  c(
    estimate = estimate,
    jackknife_bias = (n - 1) * (replicate_mean - estimate),
    jackknife_se = sqrt(
      (n - 1) / n * sum((replicates - replicate_mean)^2)
    )
  )
}

sample_values <- c(29, 79, 41, 86, 91, 5, 50, 83, 51, 42)

rbind(
  Mean = jackknife_summary(sample_values, mean),
  Median = jackknife_summary(sample_values, median)
)
```

## Smooth and non-smooth statistics

The ordinary jackknife works best for smooth statistics. Compare its standard-error estimates with a non-parametric bootstrap.

```{r}
bootstrap_standard_error <- function(x, statistic, B = 4000) {
  replicates <- replicate(
    B,
    statistic(sample(x, replace = TRUE))
  )
  sd(replicates)
}

data.frame(
  Statistic = c("Mean", "Median"),
  Jackknife_SE = c(
    jackknife_summary(sample_values, mean)["jackknife_se"],
    jackknife_summary(sample_values, median)["jackknife_se"]
  ),
  Bootstrap_SE = c(
    bootstrap_standard_error(sample_values, mean),
    bootstrap_standard_error(sample_values, median)
  )
)
```

The large disagreement for the median is a warning: deleting one observation may not describe the sampling variability of a non-smooth statistic adequately.

## BCa confidence interval

The BCa interval adjusts percentile limits for median bias and acceleration. The acceleration term is obtained from jackknife influence values.

Use the skewness of the Old Faithful eruption durations as an asymmetric statistic.

```{r}
skewness <- function(x) {
  centered <- x - mean(x)
  mean(centered^3) / mean(centered^2)^(3 / 2)
}

skewness_statistic <- function(data, indices) {
  skewness(data[indices])
}

set.seed(2026)
skewness_bootstrap <- boot(
  data = faithful$eruptions,
  statistic = skewness_statistic,
  R = 4000
)

boot.ci(
  skewness_bootstrap,
  type = c("norm", "basic", "perc", "bca")
)
```

```{r}
hist(
  skewness_bootstrap$t,
  probability = TRUE,
  breaks = 40,
  col = "grey85",
  border = "white",
  main = "Bootstrap distribution of skewness",
  xlab = "Bootstrap skewness"
)
abline(v = skewness_bootstrap$t0, col = "firebrick", lwd = 2)
```

BCa is often preferable when the bootstrap distribution is biased or skewed, but no interval method is universally best. Coverage should be checked by simulation when the population model is known.

## Bootstrap for linear regression

Fit the simple regression model

$$Y_i=\beta_0+\beta_1x_i+\varepsilon_i$$

to the built-in `cars` data.

```{r}
regression_data <- cars
regression_model <- lm(dist ~ speed, data = regression_data)

summary(regression_model)
```

```{r}
par(mfrow = c(1, 2))
plot(
  regression_data$speed,
  regression_data$dist,
  pch = 19,
  col = "grey40",
  xlab = "Speed",
  ylab = "Stopping distance",
  main = "Fitted regression"
)
abline(regression_model, col = "firebrick", lwd = 2)

plot(
  fitted(regression_model),
  residuals(regression_model),
  pch = 19,
  col = "grey40",
  xlab = "Fitted value",
  ylab = "Residual",
  main = "Residual pattern"
)
abline(h = 0, lty = 2)
par(mfrow = c(1, 1))
```

## Resampling cases

Case resampling draws $(x_i,y_i)$ pairs with replacement and refits the model. It does not condition on the observed predictor values and is less dependent on correct specification of the regression error distribution.

```{r}
bootstrap_cases <- function(data, B = 4000) {
  n <- nrow(data)
  coefficients <- matrix(
    NA_real_,
    nrow = B,
    ncol = 2,
    dimnames = list(NULL, c("Intercept", "Slope"))
  )

  for (b in seq_len(B)) {
    indices <- sample(seq_len(n), size = n, replace = TRUE)
    fit <- lm(dist ~ speed, data = data[indices, ])
    coefficients[b, ] <- coef(fit)
  }

  coefficients
}

set.seed(2026)
case_coefficients <- bootstrap_cases(regression_data)
```

## Resampling residuals

Residual resampling keeps $x_i$ fixed and generates

$$Y_i^*=\widehat Y_i+e_i^*,$$

where $e_i^*$ is sampled from the centered residuals. This method assumes that the fitted mean structure is adequate and that residuals are exchangeable, which is closely related to constant error variance.

```{r}
bootstrap_residuals <- function(model, B = 4000) {
  fitted_values <- fitted(model)
  centered_residuals <- residuals(model) - mean(residuals(model))
  original_data <- model.frame(model)
  coefficients <- matrix(
    NA_real_,
    nrow = B,
    ncol = 2,
    dimnames = list(NULL, c("Intercept", "Slope"))
  )

  for (b in seq_len(B)) {
    bootstrap_data <- original_data
    bootstrap_data$dist <- fitted_values + sample(
      centered_residuals,
      replace = TRUE
    )
    fit <- lm(dist ~ speed, data = bootstrap_data)
    coefficients[b, ] <- coef(fit)
  }

  coefficients
}

set.seed(2026)
residual_coefficients <- bootstrap_residuals(regression_model)
```

## Comparing slope uncertainty

```{r}
original_slope <- coef(regression_model)["speed"]
ols_slope_se <- summary(regression_model)$coefficients["speed", "Std. Error"]

case_interval <- quantile(
  case_coefficients[, "Slope"],
  c(0.025, 0.975)
)
residual_interval <- quantile(
  residual_coefficients[, "Slope"],
  c(0.025, 0.975)
)
ols_interval <- original_slope +
  qt(c(0.025, 0.975), df = df.residual(regression_model)) * ols_slope_se

data.frame(
  Method = c("OLS theory", "Resample cases", "Resample residuals"),
  Slope = original_slope,
  Standard_Error = c(
    ols_slope_se,
    sd(case_coefficients[, "Slope"]),
    sd(residual_coefficients[, "Slope"])
  ),
  Lower_95 = c(
    ols_interval[1],
    case_interval[1],
    residual_interval[1]
  ),
  Upper_95 = c(
    ols_interval[2],
    case_interval[2],
    residual_interval[2]
  )
)
```

```{r}
hist(
  case_coefficients[, "Slope"],
  probability = TRUE,
  breaks = 35,
  col = rgb(0.2, 0.4, 0.8, 0.45),
  border = "white",
  xlab = expression(hat(beta)[1]^"*"),
  main = "Bootstrap distributions of the slope"
)
hist(
  residual_coefficients[, "Slope"],
  probability = TRUE,
  breaks = 35,
  col = rgb(0.8, 0.2, 0.2, 0.35),
  border = "white",
  add = TRUE
)
abline(v = original_slope, lwd = 2)
legend(
  "topright",
  legend = c("Cases", "Residuals"),
  fill = c(rgb(0.2, 0.4, 0.8, 0.45), rgb(0.8, 0.2, 0.2, 0.35))
)
```

## Jackknife-after-bootstrap influence

The `boot` object retains the resampling structure required to estimate empirical influence values without rerunning an entirely new bootstrap for every deleted observation.

```{r}
slope_statistic <- function(data, indices) {
  coef(lm(dist ~ speed, data = data[indices, ]))["speed"]
}

set.seed(2026)
slope_bootstrap <- boot(
  data = regression_data,
  statistic = slope_statistic,
  R = 2000
)

slope_influence <- empinf(slope_bootstrap, type = "jack")

plot(
  slope_influence,
  type = "h",
  lwd = 2,
  xlab = "Observation",
  ylab = "Empirical influence",
  main = "Influence on the estimated slope"
)
abline(h = 0, lty = 2)
```

Large absolute influence values identify observations that deserve investigation. They are diagnostics, not automatic deletion rules.

## Key distinction

| Method | What is resampled? | Main assumption |
|----|----|----|
| Cases bootstrap | Observed $(x_i,y_i)$ pairs | Independent observational units |
| Residual bootstrap | Centered fitted residuals | Adequate mean model and exchangeable errors |

The two methods answer related questions under different assumptions. Their intervals should not be expected to agree exactly, especially when residual variability changes with the fitted value.
