---
title: "Are You Ready for a Zombie Apocalypse?"
subtitle: "Logistic Regression, Diagnostics, and Probability Prediction in R"
format:
  gfm:
    toc: true
    toc-depth: 2
    number-sections: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

This educational case study uses **logistic regression in R** to estimate the probability that a fictional observation is classified as a zombie. It began as a guided DataCamp project and was subsequently expanded with additional statistical explanations, diagnostics, model summaries, and prediction scenarios.

The workflow includes:

1. data validation and feature engineering;
2. exploratory comparisons between humans and zombies;
3. bivariate association tests;
4. logistic-regression estimation and odds-ratio interpretation;
5. model diagnostics and in-sample classification metrics;
6. probability predictions for fictional survival scenarios.

The dataset contains 200 fictional observations. The analysis is for educational and portfolio purposes.

## Setup and data loading

```{r}
#| label: setup
#| include: false

required_packages <- c("ggplot2", "gridExtra", "car", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the missing packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(ggplot2)
library(gridExtra)
library(car)
library(knitr)

theme_set(theme_minimal(base_size = 12))
options(scipen = 999)
```

```{r}
#| label: load-data

candidate_paths <- c(
  "data/zombies.csv",
  "datasets/zombies.csv",
  "zombies.csv"
)

available_paths <- candidate_paths[file.exists(candidate_paths)]

if (length(available_paths) == 0) {
  stop("zombies.csv was not found. Place it inside the data/ directory.")
}

data_path <- available_paths[1]
zombies <- read.csv(data_path, stringsAsFactors = TRUE)

cat("Dataset:", data_path, "\n")
cat("Rows:", nrow(zombies), "| Columns:", ncol(zombies), "\n")
```

The variables describe zombie status, age, sex, rurality, household size, and access to resources such as water, food, medication, tools, first-aid supplies, sanitation, clothing, and documents.

```{r}
#| label: validate-data

expected_columns <- c(
  "zombie", "age", "rurality", "household", "water", "food",
  "medication", "sanitation", "clothing", "documents"
)

missing_columns <- setdiff(expected_columns, names(zombies))
if (length(missing_columns) > 0) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

zombies$zombie <- factor(zombies$zombie, levels = c("Human", "Zombie"))

missing_summary <- data.frame(
  Variable = names(zombies),
  Missing = colSums(is.na(zombies)),
  row.names = NULL
)
missing_summary <- missing_summary[missing_summary$Missing > 0, ]

if (nrow(missing_summary) > 0) {
  kable(missing_summary, caption = "Missing values before cleaning")
} else {
  cat("No missing values detected.\n")
}
```

## Cleaning and feature engineering

In this dataset, missing clothing or documents indicate that the item is absent rather than that the response is unknown.

```{r}
#| label: feature-engineering

zombies$clothing <- as.character(zombies$clothing)
zombies$clothing[is.na(zombies$clothing)] <- "No clothing"
zombies$clothing <- factor(zombies$clothing)

zombies$documents <- as.character(zombies$documents)
zombies$documents[is.na(zombies$documents)] <- "No documents"
zombies$documents <- factor(zombies$documents)

if (any(zombies$household <= 0, na.rm = TRUE)) {
  stop("Household size must be greater than zero.")
}

zombies$water.person <- zombies$water / zombies$household

numeric_summary <- data.frame(
  Variable = c("Age", "Water per person"),
  Mean = c(mean(zombies$age), mean(zombies$water.person)),
  SD = c(sd(zombies$age), sd(zombies$water.person)),
  Median = c(median(zombies$age), median(zombies$water.person)),
  Minimum = c(min(zombies$age), min(zombies$water.person)),
  Maximum = c(max(zombies$age), max(zombies$water.person))
)

kable(numeric_summary, digits = 2, caption = "Summary of numeric variables")
```

`water.person` controls for household size and is more informative than total household water when comparing individual access to water.

## Exploring age and water availability

```{r}
#| label: fig-numeric-distributions
#| fig-cap: "Age and water availability by observed zombie status."
#| fig-width: 10
#| fig-height: 4.5

age_plot <- ggplot(zombies, aes(x = age, fill = zombie)) +
  geom_density(alpha = 0.35) +
  labs(
    title = "Age Distribution by Zombie Status",
    x = "Age",
    y = "Density",
    fill = NULL
  ) +
  theme(legend.position = "bottom")

water_plot <- ggplot(zombies, aes(x = water.person, fill = zombie)) +
  geom_density(alpha = 0.35) +
  labs(
    title = "Water per Person by Zombie Status",
    x = "Gallons of water per person",
    y = "Density",
    fill = NULL
  ) +
  theme(legend.position = "bottom")

grid.arrange(age_plot, water_plot, ncol = 2)
```

The distributions suggest that zombies tend to be older and have less water available per person. These are descriptive patterns; the plots alone do not establish causation.

## Comparing categorical characteristics

```{r}
#| label: categorical-summary

factor_predictors <- zombies[vapply(zombies, is.factor, logical(1))]
factor_predictors$zombie <- NULL

zombie_percentages <- do.call(
  rbind,
  lapply(names(factor_predictors), function(variable_name) {
    tab <- 100 * prop.table(
      table(factor_predictors[[variable_name]], zombies$zombie),
      margin = 1
    )
    data.frame(
      Variable = variable_name,
      Level = rownames(tab),
      Zombie_percentage = tab[, "Zombie"],
      row.names = NULL
    )
  })
)

kable(
  zombie_percentages,
  digits = 1,
  caption = "Observed zombie percentage within categorical predictor levels"
)
```

```{r}
#| label: fig-rurality
#| fig-cap: "Observed zombie percentage by rurality."
#| fig-width: 7
#| fig-height: 4.5

rurality_percentages <- as.data.frame(
  100 * prop.table(table(zombies$rurality, zombies$zombie), margin = 1)
)
names(rurality_percentages) <- c("Rurality", "Status", "Percentage")

ggplot(
  subset(rurality_percentages, Status == "Zombie"),
  aes(x = Rurality, y = Percentage, fill = Rurality)
) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),
    vjust = -0.3
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Observed Zombie Percentage by Rurality",
    x = NULL,
    y = "Zombie percentage"
  )
```

Urban observations have the largest percentage of zombies, while rural observations have the smallest. Food, medication, and sanitation also show visible differences, but formal tests are needed to quantify evidence for association.

## Bivariate association tests

```{r}
#| label: chi-square-tests

chi_tests <- lapply(
  factor_predictors,
  function(variable) chisq.test(table(variable, zombies$zombie))
)

chi_summary <- data.frame(
  Variable = names(chi_tests),
  Chi_squared = vapply(
    chi_tests,
    function(test) unname(test$statistic),
    numeric(1)
  ),
  Degrees_of_freedom = vapply(
    chi_tests,
    function(test) unname(test$parameter),
    numeric(1)
  ),
  P_value = vapply(chi_tests, function(test) test$p.value, numeric(1))
)

chi_summary <- chi_summary[order(chi_summary$P_value), ]
kable(chi_summary, digits = 4, caption = "Chi-square association tests")
```

```{r}
#| label: welch-tests

age_test <- t.test(age ~ zombie, data = zombies)
water_test <- t.test(water.person ~ zombie, data = zombies)

numeric_test_summary <- data.frame(
  Variable = c("Age", "Water per person"),
  T_statistic = c(unname(age_test$statistic), unname(water_test$statistic)),
  Degrees_of_freedom = c(
    unname(age_test$parameter),
    unname(water_test$parameter)
  ),
  P_value = c(age_test$p.value, water_test$p.value)
)

kable(
  numeric_test_summary,
  digits = 4,
  caption = "Welch two-sample tests"
)
```

Small p-values indicate evidence of association, not practical importance or causality. Bivariate screening is exploratory and does not replace multivariable modeling.

## Logistic regression model

The response uses `Human` as the reference level, so the model estimates the conditional probability of `Zombie`.

```{r}
#| label: logistic-model

zombie_model <- glm(
  zombie ~ age + water.person + food + rurality + medication + sanitation,
  data = zombies,
  family = binomial(link = "logit")
)

coefficient_table <- as.data.frame(summary(zombie_model)$coefficients)
coefficient_table$Term <- rownames(coefficient_table)
rownames(coefficient_table) <- NULL
coefficient_table <- coefficient_table[
  , c("Term", "Estimate", "Std. Error", "z value", "Pr(>|z|)")
]

kable(
  coefficient_table,
  digits = 4,
  caption = "Logistic-regression coefficient estimates"
)
```

```{r}
#| label: odds-ratios

model_coefficients <- summary(zombie_model)$coefficients

odds_ratio_table <- data.frame(
  Term = rownames(model_coefficients),
  Odds_ratio = exp(model_coefficients[, "Estimate"]),
  CI_2.5 = exp(
    model_coefficients[, "Estimate"] -
      1.96 * model_coefficients[, "Std. Error"]
  ),
  CI_97.5 = exp(
    model_coefficients[, "Estimate"] +
      1.96 * model_coefficients[, "Std. Error"]
  ),
  P_value = model_coefficients[, "Pr(>|z|)"],
  row.names = NULL
)

kable(
  odds_ratio_table,
  digits = 4,
  caption = "Odds ratios with 95% Wald confidence intervals"
)
```

An odds ratio above 1 indicates higher estimated odds of zombie status, while a value below 1 indicates lower estimated odds, holding the other predictors constant. Confidence intervals communicate uncertainty and should be considered together with point estimates.

```{r}
#| label: likelihood-ratio-test

null_model <- glm(
  zombie ~ 1,
  data = zombies,
  family = binomial(link = "logit")
)

likelihood_ratio_test <- anova(null_model, zombie_model, test = "Chisq")

kable(
  as.data.frame(likelihood_ratio_test),
  digits = 4,
  caption = "Likelihood-ratio test against the intercept-only model"
)
```

## Classification summary

```{r}
#| label: classification

fitted_probability <- predict(zombie_model, type = "response")

predicted_status <- factor(
  ifelse(fitted_probability >= 0.5, "Zombie", "Human"),
  levels = levels(zombies$zombie)
)

confusion_matrix <- table(
  Observed = zombies$zombie,
  Predicted = predicted_status
)

true_positive <- confusion_matrix["Zombie", "Zombie"]
true_negative <- confusion_matrix["Human", "Human"]
false_positive <- confusion_matrix["Human", "Zombie"]
false_negative <- confusion_matrix["Zombie", "Human"]
observed_binary <- as.integer(zombies$zombie == "Zombie")

classification_metrics <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Brier score", "AIC"),
  Value = c(
    (true_positive + true_negative) / sum(confusion_matrix),
    true_positive / (true_positive + false_negative),
    true_negative / (true_negative + false_positive),
    mean((observed_binary - fitted_probability)^2),
    AIC(zombie_model)
  )
)

kable(confusion_matrix, caption = "In-sample confusion matrix")
kable(classification_metrics, digits = 3, caption = "In-sample model metrics")
```

These are **in-sample** metrics: the same 200 observations were used to fit and assess the model. They describe apparent fit and may be optimistic. A train/test split or cross-validation is intentionally outside this introductory project's scope.

## Model diagnostics

```{r}
#| label: multicollinearity

vif_results <- vif(zombie_model)

if (is.matrix(vif_results)) {
  vif_table <- data.frame(Term = rownames(vif_results), vif_results, row.names = NULL)
} else {
  vif_table <- data.frame(Term = names(vif_results), VIF = unname(vif_results))
}

kable(vif_table, digits = 3, caption = "Variance-inflation diagnostics")
```

For categorical predictors, `car::vif()` reports generalized VIF values. When predictors have different degrees of freedom, the adjusted `GVIF^(1/(2*Df))` column is the appropriate comparison. Values close to 1 indicate little evidence of problematic multicollinearity.

```{r}
#| label: fig-linearity
#| fig-cap: "Informal linearity-in-the-logit diagnostics for numeric predictors."
#| fig-width: 10
#| fig-height: 4.5

bounded_probability <- pmin(pmax(fitted_probability, 1e-6), 1 - 1e-6)
zombies$logit_zombie <- qlogis(bounded_probability)

age_linearity <- ggplot(zombies, aes(x = age, y = logit_zombie)) +
  geom_point(color = "gray60", alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "orange") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black",
    linetype = "dashed"
  ) +
  labs(
    title = "Age and Estimated Log-Odds",
    x = "Age",
    y = "Estimated log-odds"
  )

water_linearity <- ggplot(zombies, aes(x = water.person, y = logit_zombie)) +
  geom_point(color = "gray60", alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "orange") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black",
    linetype = "dashed"
  ) +
  labs(
    title = "Water per Person and Estimated Log-Odds",
    x = "Gallons of water per person",
    y = "Estimated log-odds"
  )

grid.arrange(age_linearity, water_linearity, ncol = 2)
```

```{r}
#| label: fig-residuals
#| fig-cap: "Deviance residuals and influential-observation diagnostics."
#| fig-width: 10
#| fig-height: 4.5

diagnostic_data <- data.frame(
  Fitted_probability = fitted_probability,
  Deviance_residual = residuals(zombie_model, type = "deviance"),
  Cook_distance = cooks.distance(zombie_model),
  Observation = seq_len(nrow(zombies))
)

residual_plot <- ggplot(
  diagnostic_data,
  aes(x = Fitted_probability, y = Deviance_residual)
) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "orange") +
  labs(
    title = "Deviance Residuals vs Fitted Probability",
    x = "Fitted probability",
    y = "Deviance residual"
  )

influence_plot <- ggplot(
  diagnostic_data,
  aes(x = Observation, y = Cook_distance)
) +
  geom_col() +
  geom_hline(
    yintercept = 4 / nrow(zombies),
    linetype = "dashed",
    color = "orange"
  ) +
  labs(
    title = "Cook's Distance",
    x = "Observation",
    y = "Cook's distance"
  )

grid.arrange(residual_plot, influence_plot, ncol = 2)
```

These diagnostics are exploratory rather than mechanical pass/fail rules. Potentially influential observations should be investigated in context, not automatically removed.

## Scenario predictions

```{r}
#| label: scenario-predictions

scenario_data <- data.frame(
  Person = c("Dad", "Brother", "Me"),
  age = c(71, 40, 24),
  water.person = c(5, 3, 10),
  food = c("Food", "Food", "Food"),
  rurality = c("Suburban", "Urban", "Rural"),
  medication = c("Medication", "Medication", "No medication"),
  sanitation = c("Sanitation", "Sanitation", "Sanitation")
)

factor_columns <- c("food", "rurality", "medication", "sanitation")

for (column in factor_columns) {
  scenario_data[[column]] <- factor(
    scenario_data[[column]],
    levels = levels(zombies[[column]])
  )
}

scenario_data$Zombie_probability <- predict(
  zombie_model,
  newdata = scenario_data,
  type = "response"
)

scenario_results <- data.frame(
  Person = scenario_data$Person,
  Zombie_probability = scenario_data$Zombie_probability,
  Zombie_percentage = sprintf(
    "%.2f%%",
    100 * scenario_data$Zombie_probability
  )
)

kable(scenario_results, digits = 4, caption = "Fictional scenario predictions")
```

These probabilities are playful outputs from a model fitted to fictional data, not real risk estimates.

## Conclusion

This project demonstrates how exploratory analysis, hypothesis tests, and logistic regression can be combined into a coherent classification workflow. The model remains intentionally small and interpretable, with coefficients translated into odds ratios and uncertainty reported through confidence intervals.

Its principal limitations are the fictional sample, exploratory bivariate screening, in-sample evaluation, and absence of external validation. Within its educational scope, it provides a memorable introduction to logistic regression, model diagnostics, classification summaries, and probability prediction in R.
