---
title: "Bike Sharing Count Models"
subtitle: "Poisson, quasi-Poisson, and negative-binomial regression in R"
format:
  gfm:
    toc: true
    number-sections: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

This notebook models hourly Capital Bikeshare rentals using environmental and seasonal predictors. It corrects the main limitation of the original analysis: `casual` and `registered` are components of the response because

$$
cnt = casual + registered.
$$

They are therefore excluded from every predictive model.

## Setup and data validation

```{r}
required_packages <- c("ggplot2", "MASS", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop("Install the missing packages: ", paste(missing_packages, collapse = ", "))
}

library(ggplot2)
library(MASS)
library(knitr)
```

```{r}
hours <- read.csv(file.path("data", "hour.csv"))
hours$dteday <- as.Date(hours$dteday)

stopifnot(all(hours$cnt == hours$casual + hours$registered))
stopifnot(!anyNA(hours))

data.frame(
  Observations = nrow(hours),
  Variables = ncol(hours),
  StartDate = min(hours$dteday),
  EndDate = max(hours$dteday)
) |>
  kable(caption = "Dataset validation")
```

## Descriptive analysis

```{r}
ggplot(hours, aes(x = cnt)) +
  geom_histogram(bins = 60, fill = "#452D62", color = "white") +
  labs(title = "Distribution of hourly bike rentals", x = "Hourly rentals", y = "Count") +
  theme_minimal()
```

```{r}
daily_counts <- aggregate(cnt ~ dteday, data = hours, sum)

ggplot(daily_counts, aes(x = dteday, y = cnt)) +
  geom_line(alpha = 0.35, color = "gray40") +
  geom_smooth(se = FALSE, color = "#2C7FB8") +
  labs(title = "Daily bike rentals over time", x = "Date", y = "Daily rentals") +
  theme_minimal()
```

## Prepare categorical predictors

```{r}
hours$season <- factor(
  hours$season,
  levels = 1:4,
  labels = c("winter", "spring", "summer", "fall")
)
hours$yr <- factor(hours$yr, levels = c(0, 1), labels = c("2011", "2012"))
hours$mnth <- factor(hours$mnth)
hours$hr <- factor(hours$hr)
hours$holiday <- factor(hours$holiday)
hours$weekday <- factor(hours$weekday)
hours$workingday <- factor(hours$workingday)
hours$weathersit <- factor(hours$weathersit)

train_data <- subset(hours, yr == "2011")
test_data <- subset(hours, yr == "2012")

c(train = nrow(train_data), test = nrow(test_data))
```

Using 2011 for estimation and 2012 for evaluation avoids a random split that would mix neighbouring time periods. `instant`, `casual`, `registered`, and `yr` are not included in the model formula.

## Poisson and overdispersion

```{r}
base_formula <- cnt ~ season + mnth + hr + holiday + weekday +
  workingday + weathersit + atemp + hum + windspeed

poisson_model <- glm(base_formula, data = train_data, family = poisson)

poisson_dispersion <- sum(residuals(poisson_model, type = "pearson")^2) /
  df.residual(poisson_model)

poisson_dispersion
```

A dispersion value far above one indicates that the Poisson variance assumption is too restrictive.

## Quasi-Poisson

```{r}
quasipoisson_model <- glm(base_formula, data = train_data, family = quasipoisson)

quasipoisson_dispersion <- summary(quasipoisson_model)$dispersion
quasipoisson_dispersion
```

The quasi-Poisson model keeps the log-linear mean structure but adjusts coefficient standard errors for overdispersion. It does not provide a likelihood-based AIC.

## Negative-binomial models

```{r}
negative_binomial_model <- glm.nb(base_formula, data = train_data)

interaction_formula <- cnt ~ season * hr + hr * workingday +
  atemp * hum + holiday + weekday + weathersit + windspeed

negative_binomial_interactions <- glm.nb(
  interaction_formula,
  data = train_data
)
```

```{r}
head(coef(summary(negative_binomial_interactions)), 15) |>
  kable(digits = 4, caption = "First fifteen coefficients of the interaction model")
```

Because `season`, `hr`, `workingday`, and `weathersit` are factors, their coefficients and interactions represent category-specific differences rather than artificial linear changes between numeric codes.

## Temporal test-set evaluation

```{r}
poisson_prediction <- predict(poisson_model, newdata = test_data, type = "response")
nb_prediction <- predict(negative_binomial_model, newdata = test_data, type = "response")
nb_interaction_prediction <- predict(
  negative_binomial_interactions,
  newdata = test_data,
  type = "response"
)

prediction_metrics <- function(observed, predicted) {
  c(
    RMSE = sqrt(mean((observed - predicted)^2)),
    MAE = mean(abs(observed - predicted))
  )
}

test_results <- rbind(
  Poisson = prediction_metrics(test_data$cnt, poisson_prediction),
  NegativeBinomial = prediction_metrics(test_data$cnt, nb_prediction),
  NegativeBinomialInteractions = prediction_metrics(test_data$cnt, nb_interaction_prediction)
)

kable(test_results, digits = 3, caption = "Prediction error on 2012 observations")
```

```{r}
model_comparison <- data.frame(
  Model = c("Poisson", "Quasi-Poisson", "Negative binomial", "Negative binomial with interactions"),
  TrainingAIC = c(
    AIC(poisson_model),
    NA,
    AIC(negative_binomial_model),
    AIC(negative_binomial_interactions)
  )
)

kable(model_comparison, digits = 1, caption = "Likelihood-based training comparison")
```

Training AIC and future-year prediction error answer different questions, so both are reported.

```{r}
prediction_data <- data.frame(
  Date = test_data$dteday,
  Hour = test_data$hr,
  Observed = test_data$cnt,
  Predicted = nb_interaction_prediction
)

ggplot(prediction_data, aes(x = Predicted, y = Observed)) +
  geom_point(alpha = 0.2, color = "#2C7FB8") +
  geom_abline(intercept = 0, slope = 1, color = "#D7301F", linetype = 2) +
  labs(
    title = "Observed and predicted rentals in 2012",
    x = "Predicted count",
    y = "Observed count"
  ) +
  theme_minimal()
```

```{r}
first_two_weeks <- prediction_data[prediction_data$Date <= min(prediction_data$Date) + 13, ]

first_two_weeks$Time <- as.POSIXct(
  paste(first_two_weeks$Date, first_two_weeks$Hour),
  format = "%Y-%m-%d %H"
)

plot_data <- rbind(
  data.frame(Time = first_two_weeks$Time, Count = first_two_weeks$Observed, Series = "Observed"),
  data.frame(Time = first_two_weeks$Time, Count = first_two_weeks$Predicted, Series = "Predicted")
)

ggplot(plot_data, aes(x = Time, y = Count, color = Series)) +
  geom_line() +
  scale_color_manual(values = c("Observed" = "black", "Predicted" = "#A505FB")) +
  labs(title = "First two weeks of the 2012 test period", x = NULL, y = "Hourly rentals") +
  theme_minimal()
```

## Limitations

- The 2011-to-2012 split is intentionally difficult because overall demand changes over time.
- AIC is not available for quasi-Poisson models.
- Interactions increase flexibility and should be judged using the untouched future period.
- The models describe associations and predictions, not causal effects.
- `casual` and `registered` remain excluded because using components of `cnt` would make the prediction problem tautological.

